Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2020-08-30 write-up/covid-writeup.tex
Update covid-writeup.tex

The primary goal is of molecular tests is in diagnosis of active infections in the population.  Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation. Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent systemic outbreaks.  There has been work on developing faster, cheaper testing options (cite: Mina) that can be used at-home as well as group testing (citE: Mina); however, these developments are beyond the scope of this paper.  Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.

@@

To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.  These studies are useful for disease surveillance. When the number of active infections in a population are low, serological surveys are cheaper and more accurate than molecular tests and are more useful in detecting outbreaks (cite: Mina). To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.  While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.  Here we consider whether non-probability diagnostic testing can be used in estimation of prevalence of immunity.

@@

In this paper, our primary goal is to accurately estimate the active infection rate in the population over time using \emph{publicly available} test results.  Secondary goals include estimation of rates of change in these quantities and the effective reproduction number (cite:) which characterize the infectious disease spread.  A final goal is using these estimates to characterize prevalence of immunity in the general population.  As previously stated, these quantities are fundamental to public health policy and general critical information on presence and transmission of SARS-CoV-2 as well as the impact of public health  interventions.

@@

Many states provide aggregate demographic information on those who requested a test as well as those who tested positive for SARS-CoV-2. Indiana's \href{https://www.coronavirus.in.gov/2393.htm}{COVID-19 dashboard}, for instance, provides statewide demographics for age, gender, ethnicity, and race.  The demographic information is based on cumulative testing.  Daily versions cannot be easily reconstructed from available public data.  In addition, the Regenstrief Institute provides additional comorbidity information among those who tested positive for COVID-19 in the state of Indiana, including hypertension, diabetes, and chronic obstructive pulmonary disease (COPD). These comprise the main set of covariate information available from public sources for Indiana. Similar demographic information is likely available for other states.co
% https://www.regenstrief.org/covid-dashboard/
%

@@

We focus on US COVID-19 testing.

The typical testing

However, upon investi
%https://www.coronavirus.in.gov/
As of today, access to the history of testing requirements remains quite difficult.  Early during the coronavirus pandemic, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); however, as testing capacity expanded these restrictions were gradually lifted.  Such restrictions

Given the difficult, we focus on reconstructing this history for the state of Indiana. As of May 12th, 2020, the Indiana State Department of Health laboratory was testing high-risk individuals for COVID-19. Any symptomatic individual could receive a tested.  Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness.
% https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities.
% https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus.
% https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/

>>>>

The primary goal is of molecular tests is in diagnosis of active infections in the population.  Such diagnostic tests generate important information about the presence of SARS-Cov-2 in the population, and help scientists and policy-makers understand patterns of transmission and propagation. Rapid and frequent molecular testing is cited as a key component (citEe: OECD and others) in effective strategies to identify active infections and prevent systemic outbreaks.  There has been work on developing faster, cheaper testing options (cite: Mina) that can be used at-home as well as group testing (citE: Mina); however, these developments are beyond the scope of this paper.
% Here we consider the current system of testing which involves both self-selection into testing and historical limits on who was able to receive diagnostic tests due to limited testing capacity.

@@

To better estimate SARS-CoV-2 immunity in a population, seroprevalence studies that generate a probabilistic population sample and perform serological tests on the sample can be collected.  These studies are useful for disease surveillance. When the number of active infections in a population are low, serological surveys are cheaper and more accurate than molecular tests and are more useful in detecting outbreaks (cite: Mina). To date, the CDC has conducted ten large-scale geographic serological surveys with three rounds.  While population-based sampling strategies provide a more representative nationwide sample, they are very time intensive and expensive.

@@

In this paper, our primary goal is accurate estimates of the population-level active infection rates over time using \emph{publicly available} testing data.  Secondary goals include estimation of rates of change in these quantities and the effective reproduction number (cite:) which characterize the infectious disease spread.  A final goal is using these estimates to characterize prevalence of immunity in the general population.  As previously stated, these quantities are fundamental to public health policy and general critical information on presence and transmission of SARS-CoV-2 as well as the impact of public health  interventions.

@@

Many states also provide aggregate demographic information on those who requested a test as well as those who tested positive for SARS-CoV-2. Indiana's \href{https://www.coronavirus.in.gov/2393.htm}{COVID-19 dashboard}, for instance, provides statewide demographics for age, gender, ethnicity, and race.  The demographic information is based on cumulative testing.  In addition, the Regenstrief Institute provides additional comorbidity information among those who tested positive for COVID-19 in the state of Indiana, including hypertension, diabetes, and chronic obstructive pulmonary disease (COPD) summary statistics. These comprise the main set of covariate information available from public sources for Indiana. Similar demographic information is likely available for other states.
% https://www.regenstrief.org/covid-dashboard/
%

@@

Due to limited testing capacity, many US states have instituted restrictions on who can request a molecular test at a given time.
%https://www.coronavirus.in.gov/
As of August 30th, 2020, a repository of the history of testing requirements does not exist.  Early on, limited testing meant that only those at high risk were allowed to be tested (i.e., symptomatic or high risk populations); as testing capacity expanded, these restrictions were gradually lifted.

Here, we focus on reconstructing the testing restriction history for the state of Indiana. As of May 12th, 2020, the Indiana State Department of Health (ISDH) laboratory was testing high-risk individuals for COVID-19. Any symptomatic individual could receive a tested.  Moreover there were key factors for people who can be tested regardless of symptom status: over the age of 65, diabetic, high-blood pressure, obesity, pregnancy, and minority populations at greater risk of severe illness.
% https://www.indystar.com/story/news/health/2020/05/12/coronavirus-testing-indiana-who-should-get-tested/3110592001/
On April 28th, 2020, the criteria expanded to include any Indiana resident with virus symptoms, people in close contact with those who have tested positive and residents of congregant communities.
% https://www.wishtv.com/news/medical/indiana-opens-up-covid-19-testing-to-more-hoosiers/
On June 15, 2020, ISDH lifted all restrictions and testing expanded to include anyone in Indiana who wants to be tested for coronavirus.
% https://www.indystar.com/story/news/health/2020/06/12/indiana-says-anyone-who-wants-coronavirus-test-can-get-one/3179151001/
